---
title: "Germination Data Analysis Student Version"
author: "YOUR NAME"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
    html-table-processing: none
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## Learning Objectives

- Learn how to implement tidyverse functions to analyze and interpret real-world ecological germination data.

- Create summary statistics + visualizations to compare germination rates across treatments, species, concentrations, and environmental conditions.

- Practice writing reusable R functions using tidy evaluation techniques such as the embrace operator (`{{ }}`).

- Use iteration tools from the `purrr` package to efficiently repeat analyses across multiple treatment groups or variables.

- Explore distributions + simulation methods to better understand variability in germination outcomes.

- Practice writing reproducible, well-documented, + readable R code for ecological data analysis.

## Dataset Description

The dataset used in this lab contains germination data collected during a 2025 summer ecology study from Carson investigating the effects of fire retardants on grassland species. Specifically, the study examined how two fire retardants, Phos-Chek and Clore, influenced the germination of Purple Needle Grass (PNG) and Italian Ryegrass (RYE).

For each observation, the number of seeds germinated out of 10 total seeds was recorded weekly over a four-week period. The dataset includes the following variables:

- `week` – week of observation during the experiment
- `species` – plant species being tested
- `substrate` – substrate or soil type used
- `treatment_type` – fire retardant treatment applied
- `concentration` – concentration level of treatment
- `dish_number` – petri dish identification number
- `seeds_germinated_out_of_10` – number of seeds germinated out of 10 total seeds

This dataset contains a large number of observations and multiple treatment combinations, that allows for practice with data visualization, summary statistics, iteration, functions, and simulation techniques.

## Download the Data

Download the dataset for this lab: [Germination_Data_Entry_1.xlsx](Germination_Data_Entry_1.xlsx)

Save this Excel sheet in your project folder before running the code in this lab.

## Import Libraries

::: {.callout-note}
## Note

If you have not previously installed these packages, run:

```r
install.packages("readxl")
install.packages("tidyverse")
install.packages("janitor")
install.packages("knitr")
install.packages("gt")
```

in the Console once. After the packages are installed, you can load them in future sessions using `library()`.
:::

```{r}
#| message: false
# Import your libraries here
```

## Read Data

::: {.callout-warning}
## Warning

If you change the name of the dataset file, be sure to update the filename used in `read_excel()` below.
:::

```{r}
#| message: false
germination <- read_excel("Germination_Data_Entry_1.xlsx")
germination <- clean_names(germination)
```

## Preview the Data

**Question 0: Let's first poke around and get a sense of the dataset. What is the number of observations and variables? Are there any missing values?**

::: {.callout-tip}
## Tip

You can use `glimpse()` and `colSums(is.na())` to answer the question above!
:::

Write a short description here.

```{r}
# Add your code here
```

## Questions

**Question 1: Which treatment type had the highest average germination percentage for each species? Create a summary table and display the relationship between treatment type and average germination percentage using a bar plot.**

```{r}
# Add your table code here
```

```{r}
# Add your bar plot code here
```

Write a short description here.

**Question 2: Create a faceted line graph detailing the frequencies for seeds germinated between different concentrations. Lines should be separated by the treatment and it should be faceted by the weeks.** **Hint: There are 4 dishes for each concentration so the frequency should be out of 40.**

```{r}
# Add your graph code here
```

**Question 3: Create a table using gt() that displays the mean germination for each species and fire retardant from week 4.** **Hint: there should be 3 columns and 4 rows.**

```{r}
# Add your table code here
```

**Question 4: Write a reusable function called `germination_success()` that calculates the percentage of observations where at least half of the seeds germinated. For this question, define a successful observation as one where at least 5 out of 10 seeds germinated. Use your function to calculate germination success percentages by species.**

Your function should:

- use a grouping variable chosen by the user
- use the embrace operator (`{{ }}`)
- return the success percentage for each group

```{r}
germination_success <- function(data, var){
  # add your function code here
}
```

```{r}
# Add your function test code here
```

Write a short description here.

**Question 5: Create a new variable called `germination_levels`, then create a summary table showing the number of observations in each germination level for each species.**

Classify the observations as:

- Low: 0–3 seeds germinated
- Moderate: 4–7 seeds germinated
- High: 8–10 seeds germinated

```{r}
# Add your table code here
```

Write a short description here.

**Question 6: Use `map()` to count how many Low, Moderate, and High germination observations occurred in each week. Which week had the strongest germination performance?**

::: {.callout-note}
## Note

There are multiple ways to solve this problem. While this solution uses `map()` to practice iteration, you can achieve the same result using `group_by()` and `summarize()`. You may try exploring an alternative solution on your own.
:::

```{r}
# Add your table code here
```

Write a short description here.

**Question 7: Gather the mean and standard deviation of the seeds germinated for the treatment type Clore. With those, create a simulation 1000 samples and graph the distribution. Does this graph makes sense to create, why or why not?**

```{r}
# Add your table code here
```

```{r}
set.seed(3452)

# Add your distribution code here
```

**Question 8: For each of the week, substrate and treatment find the frequency of seeds germinated/10, and identify the highest frequency for each week and make sure to format the the table with a caption for the table. Hint: use the same frequency calculation you did for question 2.**

```{r}
# Add your table code here

```

**Question 9: Use functions from the purrr package to calculate germination summaries separately for each species. Hint: create a summary function, summarize the dataset using nest(), apply a summary function to each subset using a map function, then combine the results into a single tidy table.**

```{r}
#Summary function:

species_summary <- function(data) {
  # Add your function code here
}

# Nest by species and apply summary function:

# Add your table code here
```

Write a short description of your table here.

**Question 10: Building on the code from question 9, build a table that summarizes the data by BOTH species and substrate. Hint: use nest() and unnest().**

```{r}
# Add your table code here
```

Write a short description of your table here.
