---
title: "Group W Project: Germination Data Analysis"
author: "Aarav, Carson, and Jared"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
    html-table-processing: none
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## Learning Objectives

- Learn how to implement tidyverse functions using real-world ecological datasets and interpret their impact on germination patterns.

- Create summary statistics and visualizations to compare germination rates across treatments, species, and environmental conditions.

- Practice writing reusable R functions using tidy evaluation techniques such as the embrace operator (`{{ }}`).

- Use iteration tools from the `purrr` package to efficiently repeat analyses across multiple treatment groups or variables.

- Explore distributions, sampling, and simulation methods to better understand variability in germination percentages.

- Practice writing reproducible and well-documented R code for ecological and plant biology data analysis.

## Dataset Description

The dataset used in this lab contains germination data collected during a 2025 summer ecology study from Carson investigating the effects of fire retardants on grassland species. Specifically, the study examined how two fire retardants, Phos-Chek and Clore, influenced the germination of Purple Needle Grass (PNG) and Italian Ryegrass (RYE).

For each observation, the number of seeds germinated out of 10 total seeds was recorded weekly over a four-week period. The dataset includes the following variables:

- `week` – week of observation during the experiment
- `species` – plant species being tested
- `substrate` – substrate or soil type used
- `treatment_type` – fire retardant treatment applied
- `concentration` – concentration level of treatment
- `dish_number` – petri dish identification number
- `seeds_germinated_out_of_10` – number of seeds germinated out of 10 total seeds

This dataset contains a large number of observations and multiple treatment combinations, that allows for practice with data visualization, summary statistics, iteration, functions, and simulation techniques.

## Import Libraries

```{r}
#| message: false

library(readxl)
library(tidyverse)
library(janitor)
library(knitr)
library(gt)
```

## Read Data

```{r}
germination <- read_excel("Germination_Data_Entry_1.xlsx")
germination <- clean_names(germination)
```

## Preview the Data

```{r}
glimpse(germination)
colSums(is.na(germination))
```

## Questions

**Question 1: Which treatment type had the highest average germination percentage for each species? Create a summary table and display the relationship between treatment type and average germination percentage using a bar plot.**

```{r}
treatment_summary <- germination |>
  group_by(species, treatment_type) |>
  summarize(
    average_germination_percent =
    mean(seeds_germinated_out_of_10 / 10 * 100),
    .groups = "drop"
  )

treatment_summary |>
  kable()
```

```{r}
ggplot(treatment_summary,
       aes(x = treatment_type,
           y = average_germination_percent,
           fill = treatment_type)) +
  geom_col() +
  facet_wrap(~ species) +
  labs(
    title = "Average Germination Percentage by Treatment",
    x = "Treatment Type",
    y = "Average Germination Percentage"
  )
```
For PNG, the Clore treatment had a slightly higher average germination percentage than Phos. In contrast, RYE had a much higher germination percentage under the Phos treatment. The bar plot makes these differences easier to visualize by clearly showing how the treatments affected each species differently.

**Question 2: Create a faceted line graph detailing the frequencies for seeds germinated between different concentrations. Lines should be separated by the treatment and it should be faceted by the weeks**
**Hint: There are 4 dishes for each concentration so the frequency should be out of 40.**
```{r}
germination |> 
  mutate(week=fct_recode(as.character(week),"Week 1"="1",
                         "Week 2"="2",
                         "Week 3"="3",
                         "Week 4"="4")) |> 
  group_by(week,treatment_type,concentration) |> 
  summarise(freq=sum(seeds_germinated_out_of_10)/40,
            .groups="drop") |> 
  ggplot(mapping=aes(x=concentration,
                     y=freq,
                     color=treatment_type))+
  geom_line()+
  facet_wrap(~week)+
  labs(title="Seeds Germination Rate Among Different Concentrations",
       subtitle="Rate",
       x="Conentration",
       y="",
       color="Treatment")
```

**Question 3: Create a table using gt() that displays the mean germination for each species and fire retardant from week 4.**
**Hint: there should be 3 columns and 4 rows.**
```{r}
germination |>
  
  filter(week == 4) |>
  
  group_by(species, treatment_type) |>
  
  summarize(
    mean_germination = mean(seeds_germinated_out_of_10, na.rm = TRUE),
    .groups = "drop") |>
  
  gt() |>
  tab_header(
    title = "Mean Germination at Week 4",
    subtitle = "Grouped by Species and Fire Retardant Treatment") |>
  
   fmt_number(
    columns = mean_germination,
    decimals = 3) |>
  
  cols_label(
    species = "Species",
    treatment_type = "Fire Retardant",
    mean_germination = "Mean Germination (out of 10)")
```

**Question 4: Write a reusable function called `germination_success()` that calculates the percentage of observations where at least half of the seeds germinated. For this question, define a successful observation as one where at least 5 out of 10 seeds germinated. Use your function to calculate germination success percentages by species.** 

Your function should:

- use a grouping variable chosen by the user
- use the embrace operator (`{{ }}`)
- return the success percentage for each group

```{r}
germination_success <- function(data, var){
  data |>
    group_by({{ var }}) |>
    summarize(
      success_percent =
        round(mean(seeds_germinated_out_of_10 >= 5) * 100, 2)
    )
}
```

```{r}
germination_success(germination, species) |>
  kable(caption = "Germination Success Percentage by Species")
```
The table shows the percentage of observations where at least 5 out of 10 seeds successfully germinated for each species. These results suggest that RYE generally germinated more successfully than PNG under the treatment conditions included in this dataset.

**Question 5: Create a new variable called `germination_level`, then create a summary table showing the number of observations in each germination level for each species.**

Classify the observations as:

- Low: 0–3 seeds germinated
- Moderate: 4–7 seeds germinated
- High: 8–10 seeds germinated

```{r}
germination_levels <- germination |>
  mutate(
    germination_level = factor(
      case_when(
        seeds_germinated_out_of_10 <= 3 ~ "Low",
        seeds_germinated_out_of_10 <= 7 ~ "Moderate",
        seeds_germinated_out_of_10 <= 10 ~ "High"),
      levels = c("Low", "Moderate", "High")
    )
  )
germination_levels |>
  group_by(species, germination_level) |>
  summarize(
    observations = n(),
    .groups = "drop"
  ) |>
  kable(caption = "Observations by Germination Level and Species")
```
Most observations for both species were classified in the Low germination category, meaning that 0–3 seeds germinated in the majority of cases. These results also suggest that RYE generally had stronger germination outcomes than PNG under the treatment conditions included in this dataset.

**Question 6: Use `map()` to count how many Low, Moderate, and High germination observations occurred in each week. Which week had the strongest germination performance?**

```{r}
week_levels <- map(
  unique(germination_levels$week),
  ~ germination_levels |>
    filter(week == .x) |>
    count(germination_level,
          name = "observations",
          .drop = FALSE) |>
    mutate(week = .x)
)

bind_rows(week_levels) |>
  kable(caption = "Germination Levels by Week")
```
These results suggest that germination outcomes improved over time for some observations, with Week 3 showing the strongest overall germination performance since it had the most High observations with 5.

**Question 7: Gather the mean and standard deviation of the seeds germinated for the treatment type Clore. With those create a simulation 1000 samples and graph that distribution. Does this graph makes sense to create why or why not?**


```{r}
germination|> 
  filter(treatment_type=="Clore") |> 
  group_by(treatment_type) |> 
  summarize(avg=mean(seeds_germinated_out_of_10),
            sd=sd(seeds_germinated_out_of_10))
```

```{r}
set.seed(3452)
sim=data.frame(val=rnorm(c(1:1000),mean=1.34,sd=1.84))

sim |> 
  ggplot(mapping = aes(x=val))+
  geom_histogram()

```
Based on the graph that was created and from the values of the mean and standard deviation, the graph wouldn't make sense for it to be made as a normal distribution. This is because the standard deviation is larger than mean and that could be based on varying some of the values were originally. So when this simulation was created using the values we attained, it doesn't look symmetric and instead is a bit off. 

**Question 8: For each of the week and substrate, identify the number of seeds germinated/10 that occured the most make sure to format the the table with a title and good column labels**
```{r}

```