---
title: "Stat331GroupWProject"
format: html
editor: visual
---

## Learning Objectives

- Learn how to implement tidyverse functions using real-world ecological datasets and interpret their impact on germination patterns.

- Create summary statistics and visualizations to compare germination rates across treatments, species, and environmental conditions.

- Practice writing reusable R functions using tidy evaluation techniques such as the embrace operator (`{{ }}`).

- Use iteration tools from the `purrr` package to efficiently repeat analyses across multiple treatment groups or variables.

- Explore distributions, sampling, and simulation methods to better understand variability in germination percentages.

- Practice writing reproducible and well-documented R code for ecological and plant biology data analysis.

## Dataset Description

The dataset used in this lab contains germination data collected during a 2025 summer ecology study from Carson investigating the effects of fire retardants on grassland species. Specifically, the study examined how two fire retardants, Phos-Chek and Clore, influenced the germination of Purple Needle Grass (PNG) and Italian Ryegrass (RYE).

For each observation, the number of seeds germinated out of 10 total seeds was recorded weekly over a four-week period. The dataset includes the following variables:

- `week` – week of observation during the experiment
- `species` – plant species being tested
- `substrate` – substrate or soil type used
- `treatment_type` – fire retardant treatment applied
- `concentration` – concentration level of treatment
- `dish_number` – petri dish identification number
- `seeds_germinated_out_of_10` – number of seeds germinated out of 10 total seeds

This dataset contains a large number of observations and multiple treatment combinations, that allows for practice with data visualization, summary statistics, iteration, functions, and simulation techniques.

## Import Libraries

```{r}
#| message: false

library(readxl)
library(tidyverse)
library(janitor)
library(knitr)
```

## Read Data

```{r}
germination <- read_excel("Germination_Data_Entry_1.xlsx")
germination <- clean_names(germination)
```

## Preview the Data

```{r}
glimpse(germination)
colSums(is.na(germination))
```

## Questions

**Question 1: Create a summary table showing the average germination percentage for each concentration level, treatment type, and species combination. Based on the table, describe any patterns you observe between concentration levels and germination percentages.**

```{r}
concentration_summary <- germination |>
  group_by(species, treatment_type, concentration) |>
  summarize(
    average_germination_percent = mean(seeds_germinated_out_of_10 / 10 * 100),
    .groups = "drop"
  ) |>
  kable()
  
concentration_summary
```

Description: 

Overall, RYE usually had higher germination percentages than PNG for most treatments.

For the Clore treatment, germination percentages generally became lower as concentration levels increased for both
species. For RYE, the Phos treatment often had higher germination percentages than Clore at several concentration levels.

In general, higher concentration levels were usually linked to lower germination percentages, although this pattern was
not exactly the same in every case, but there appears to be a noticeable trend.

These results suggest that both treatment type and concentration level may affect germination success, and that the
treatments may affect each species differently.


**Question 2: Create a faceted line graph detailing the frequencies for seeds germinated between different concentrations. Lines should be separated by the treatment and it should be faceted by the weeks**
**Hint: There are 4 dishes for each concentration so the frequency should be out of 40**
```{r}
germination |> 
  mutate(week=fct_recode(as.character(week),"Week 1"="1",
                         "Week 2"="2",
                         "Week 3"="3",
                         "Week 4"="4")) |> 
  group_by(week,treatment_type,concentration) |> 
  summarise(freq=sum(seeds_germinated_out_of_10)/40,
            .groups="drop") |> 
  ggplot(mapping=aes(x=concentration,
                     y=freq,
                     color=treatment_type))+
  geom_line()+
  facet_wrap(~week)+
  labs(title="Seeds Germination Rate Among Different Concentrations",
       subtitle="Rate",
       x="Conentration",
       y="",
       color="Treatment")
```