---
title: "Stat331GroupWProject"
format: html
editor: visual
---

## Learning Objectives

- Learn how to implement tidyverse functions using real-world ecological datasets and interpret their impact on germination patterns.

- Create summary statistics and visualizations to compare germination rates across treatments, species, and environmental conditions.

- Practice writing reusable R functions using tidy evaluation techniques such as the embrace operator (`{{ }}`).

- Use iteration tools from the `purrr` package to efficiently repeat analyses across multiple treatment groups or variables.

- Explore distributions, sampling, and simulation methods to better understand variability in germination percentages.

- Practice writing reproducible and well-documented R code for ecological and plant biology data analysis.

## Dataset Description

The dataset used in this lab contains germination data collected during a summer ecology study investigating the effects of fire retardants on grassland species. Specifically, the study examined how two fire retardants, Phos-Chek and Clore, influenced the germination of Purple Needle Grass (PNG) and Italian Ryegrass (RYE).

For each observation, the number of seeds germinated out of 10 total seeds was recorded weekly over a four-week period. The dataset includes:

- species
- substrate type
- treatment type
- concentration
- dish number
- weekly germination counts

This dataset contains a large number of observations and multiple treatment combinations, that allows for practice with data visualization, summary statistics, iteration, functions, and simulation techniques using real-world ecological data.

## Read Data

```{r}
#| message: false
library(readxl)
library(tidyverse)
library(janitor)

germination <- read_excel("Germination_Data_Entry_1.xlsx")
germination <- clean_names(germination)
```

## Preview the Data
```{r}
glimpse(germination)

head(germination)
```

